---
title: "**MATH 3190 Homework 6**"
subtitle: |
  Due March 19, 2026\
  Focus: Notes 7 Part 2\
  Total Points: 90 + 10 (Self-Critique Comment and Uploading to GitHub) = 100
format: 
  html:
    embed-resources: true # Keep this so the html file is easily shareable.
urlcolor: blue
execute: # Will set these as the default for all code chunks
  eval: true
  echo: true
editor_options: 
  chunk_output_type: console
---

Note that your homework should be completed in Quarto. You can just add your answers and code to this document in the appropriate part. The file should be rendered as an html document or pdf document (html is quicker to render and you will need an installation of LaTeX to render as a pdf). You will "turn in" this homework by uploading the files, the .qmd and output file, to your GitHub `Math_3190_Assignment` repository in the `Homework/Homework_6` directory. Make sure all supporting files (if there are any) are in this directory as well. You should use `git` and `Unix` commands in a terminal for this (i.e. don't manually upload files to GitHub). In addition, you will submit just your html or pdf file on Canvas to unlock the solutions.

Don't forget to also put a [self-critique comment]{.underline} on Canvas indicating what you did correctly and incorrectly and include any questions you may have about the assignment after submitting the assignment and thoroughly looking through the solutions.

# Problem 1 (28 points)

An automobile consulting company wants to understand the factors on which the pricing of cars depends. The dataset `car_price_prediction` in the `math3190package` (and also found in the GitHub data folder) has information on the sales of 4340 vehicles.

```{r}
library("tidyverse")
library("math3190package")

data(car_price_prediction)
```

### Part a (3 points)

Create new variables in the dataset by taking the natural log of selling price, year, and km_driven. This will help linearize the association between variables. Then fit a linear model for predicting the log of selling price using all other variables except "name", "year", "selling_price", and "km_driven" (the logged variables should be used instread of those last three). Also, there is only one vehicle with a fuel type of "Electric". Remove that vehicle from the dataset.

```{r}
x = mutate(car_price_prediction, log_sell = log(selling_price))
y = mutate(x, log_year = log(year))
car_predict = mutate(y, log_km = log(km_driven))

electric_car = which(car_predict$fuel == "Electric")
car_predict = car_predict[-electric_car, ]

model = lm(log_sell ~ fuel + seller_type + transmission + owner + log_year + log_km, data = car_predict)
se = summary(model)$coefficients[-1, 2]
```

### Part b (5 points)

Now fit some LASSO models for predicting log selling price using all but the "name" variable. Try the following values for the regularization parameter: $\lambda=0,\ 0.01,\ 0.1,$ and $1$ and comment on how the coefficients of the model change.

Note: when $\lambda=0$, the LASSO coefficients should equal the OLS model coefficients. However, they will actually be a bit off here. That is because the `glmnet()` function has a `thresh` argument that sets a threshold for convergence. It is, by default, set to `1e-07`. To make the parameters match, you can change that to `thresh = 1e-14` instead. This is not necessary, though.

```{r}
library(glmnet)

x = model.matrix(log_sell ~ fuel + seller_type + transmission + owner + log_year + log_km, data = car_predict)[, -1]
y = car_predict$log_sell

lasso_0 = glmnet(x, y, alpha = 1, lambda = 0)
lasso_1 = glmnet(x, y, alpha = 1, lambda = 0.01)
lasso_2 = glmnet(x, y, alpha = 1, lambda = 0.1)
lasso_3 = glmnet(x, y, alpha = 1, lambda = 1)

coef(lasso_0)
coef(lasso_1)
coef(lasso_2)
coef(lasso_3)

# lasso_0 has every single coefficient, most of which are of to the -1 or -2 power
# lasso_1 has smaller coefficients, with some even reducing to 0
# lasso_2 has even more coefficients reduced to 0
# lasso_3 has every coefficient reduced to 0
```

### Part c (4 points)

Now fit some ridge regression models for predicting log selling price using all but name. Try the following values for the regularization parameter: $\lambda=0,\ 0.01,\ 0.1,$ and $1$ and comment on how the coefficients of the model change.

```{r}
library(glmnet)

ridge_0 = glmnet(x, y, alpha = 0, lambda = 0)
ridge_1 = glmnet(x, y, alpha = 0, lambda = 0.01)
ridge_2 = glmnet(x, y, alpha = 0, lambda = 0.1)
ridge_3 = glmnet(x, y, alpha = 0, lambda = 1)

coef(ridge_0)
coef(ridge_1)
coef(ridge_2)
coef(ridge_3)

# Each lambda level doesn't have any zeros. Some variables decrease as lambda increases, some increase as lambda increases, and some do neither. 
```

### Part d (4 points)

Now fit some elastic net regression models for predicting log price using all but name. Try the following values for the regularization parameter: $\lambda=0,\ 0.01,\ 0.1,$ and $1$ for $\alpha=1/3$ and then comment on how the coefficients of the model change for each $\alpha$ from parts (b), (c), and (d).

```{r}
net_0 = glmnet(x, y, alpha = 1/3, lambda = 0)
net_1 = glmnet(x, y, alpha = 1/3, lambda = 0.01)
net_2 = glmnet(x, y, alpha = 1/3, lambda = 0.1)
net_3 = glmnet(x, y, alpha = 1/3, lambda = 1)

coef(net_0)
coef(net_1)
coef(net_2)
coef(net_3)

# This model has more zeros as lambda increases, similar to LASSO. However, this time, not all the variables decrease to zero; log_year is still there, unlike in LASSO. In the LASSO regression, the zeros appeared in a faster rate as lambda increased compared to elastic net. For example, in the elastic net, no zeros appear, similar to the ridge regression model. 
```

### Part e (4 points)

Use the `cv.glmnet()` to find the "optimal" $\lambda$ value (let's use the "lambda.1se") for LASSO ($\alpha=1$), ridge ($\alpha=0$), and elastic net (let's again use $\alpha = 1/3$) and then fit models using the `glmnet()` function for all three using their respective "best" $\lambda$. **Set a seed** before running the `cv.glmnet()` each time.

Compare each model's variables and coefficients.

```{r}
set.seed(39)
cv_lasso = cv.glmnet(x, y, alpha = 1)
print(cv_lasso$lambda.min)

cv_ridge = cv.glmnet(x, y, alpha = 0)
print(cv_ridge$lambda.min)

cv_net = cv.glmnet(x, y, alpha = 1/3)
print(cv_net$lambda.min)

lasso = glmnet(x, y, alpha = 1, lambda = 0.001825386)
ridge = glmnet(x, y, alpha = 0, lambda = 0.05839897)
net = glmnet(x, y, alpha = 1/3, lambda = 0.004142512)

coef(lasso)
coef(ridge)
coef(net)

# For the lasso and net models, the fuelPetrol variable has been zeroed out. These are the only instances of a zero occurring in these models. Otherwise, the coefficients are similar. 
```

### Part f (8 points)

Use bootstrapping with at least 1000 samples to estimate the standard errors of the coefficients of the ridge regression model. For each bootstrap sample, run the `cv.glmnet()` function to find the best $\lambda$ and fit a model using that optimized $\lambda$. Use the "lambda.1se" for this. This can take a few minutes to run. Then compare these bootstrapped standard errors to the standard errors for the OLS model you fit in part a. Are they larger or smaller? Does this make sense?

```{r}
#| cache: true
set.seed(0831)
bootstrap_iterations = 1000
n = nrow(x)
p = ncol(x)

bootstrap_coefficients = matrix(NA, nrow = bootstrap_iterations, ncol = p)

for (b in 1:bootstrap_iterations) {
  index = sample(1:n, size = n, replace = TRUE)
  x_boot = x[index, ]
  y_boot = y[index]
  cv_fit = cv.glmnet(x_boot, y_boot, alpha = 0)
  best_lambda = cv_fit$lambda.1se
  ridge_fit = glmnet(x_boot, y_boot, alpha = 0, lambda = best_lambda)
  bootstrap_coefficients[b, ] = as.vector(coef(ridge_fit))[-1]
}

boot_se = apply(bootstrap_coefficients, 2, sd)
boot_se
comparison = data.frame(
  OLS_SE = se,
  Ridge_SE = boot_se
)

# The ridge regression values are smaller except for log_year and transmissionManual. This makes sense, since ridge regression introduces a penalty that moves the errors closer to zero. This leads to smaller errors with a bit of bias. 
```

# Problem 2 (21 points)

In 1990, sixty districts in California were randomly selected with the goal of predicting the median house value for the district based on the district location (longitude and latitude), the median house age in the block group, the total number of rooms in the homes, the total number of bedrooms, the population in the district, the number of households, and the median income (in \$10,000). The data are already in the .qmd file.

```{r}
#| label: house_data
#| echo: false
library(tidyverse) |> suppressPackageStartupMessages()
housing_train <- tibble(
  house_value = c(156100, 145300, 150000, 50600, 176300, 178500, 123800,
                  173900, 422700, 164400, 42100, 157800, 162900, 90500,
                  191800, 128100, 109700, 206700, 132200, 248200, 336900,
                  58000, 271500, 259500, 69400, 132700, 461200, 245000,
                  218000, 311700, 361600, 250000, 113900, 103600, 81900,
                  351900, 114100, 500001, 153300, 466700),
  longitude = c(-121.08, -121.27, -117.7, -119.29, -118.09, -117.67, 
                -120.76, -118.16, -118.39, -123.1, -117.66, -118.16, 
                -120.98, -120.99, -122.69, -118.18, -122.27, -122.11, 
                -121.99, -118.02, -118.02, -120.94, -118.6, -121.95, 
                -120.62, -118.18, -122.13, -117.93, -121.85, -118.49, 
                -122.44, -121.32, -117.07, -121.83, -121.43, -122.11, 
                -118.27, -118.4, -120.28, -117.16),
  latitude = c(38.95, 38.7, 34.06, 36.34, 33.9, 34.02, 38.47, 34.02,
               33.89, 38.79, 35.63, 33.97, 38.99, 37.67, 38.35, 33.91,
               37.83, 37.99, 38.34, 33.74, 33.73, 40.14, 34.26, 37.96, 
               36.99, 33.8, 37.4, 33.94, 36.6, 34.18, 37.76, 37.67, 32.74, 
               38, 38.53, 37.41, 34.01, 34.41, 37.9, 32.74),
  age = c(18, 16, 25, 28, 33, 16, 17, 44, 38, 20, 33, 23, 17, 28, 16, 41, 
          51, 16, 16, 26, 24, 31, 18, 18, 32, 42, 29, 30, 21, 31, 52, 21, 
          38, 15, 36, 27, 47, 22, 17, 43),
  rooms = c(1931, 3747, 2054, 1440, 3326, 3042, 1521, 1218, 1851, 3109, 
            2579, 1516, 3403, 1768, 1689, 1260, 2665, 3913, 1470, 3842, 
            6393, 3127, 6154, 2739, 2455, 2301, 6027, 2658, 2381, 3073, 
            2959, 1494, 1901, 6365, 2430, 5110, 921, 4443, 1047, 1437),
  bedrooms = c(380, 586, 609, 431, 720, 524, 309, 374, 332, 712, 564, 
               457, 661, 423, 254, 299, 574, 710, 261, 609, 1141, 664, 
               1070, 393, 508, 621, 1195, 382, 701, 674, 683, 271, 392, 
               1646, 426, 1599, 264, 560, 212, 406),
  population = c(1271, 1817, 2271, 2178, 2533, 1516, 607, 1175, 750, 1643,
                 1155, 1977, 1540, 1066, 921, 1535, 1258, 1782, 748, 1961,
                 2743, 1345, 3010, 1072, 1344, 2114, 2687, 1135, 1264, 1486,
                 1145, 781, 1099, 3838, 1199, 2764, 881, 1573, 530, 692),
  households = c(377, 590, 564, 440, 689, 475, 240, 342, 314, 638, 431, 435,
                 622, 392, 270, 322, 536, 676, 256, 595, 1057, 580, 1034, 374,
                 492, 561, 1171, 392, 659, 684, 666, 255, 406, 1458, 437,
                 1482, 221, 496, 196, 379)
)

housing_test <- tibble(
  house_value = c(214900, 44000, 194400, 128700, 435000, 179400, 129300,
                  312300, 76300, 225800, 295500, 162500, 340800, 93800,
                  155000, 283200, 133400, 248200, 265000, 57500),
  longitude = c(-119.25, -121.93, -122.31, -121.27, -118.42, -118.71, 
                -117.31, -122.04, -119.69, -117.88, -117.25, -121.48, 
                -122.49, -121.49, -116.18, -119.19, -118.23, -118.35, 
                -118.35, -124.15),
  latitude = c(34.26, 41.86, 38.27, 38.67, 33.85, 34.27, 34.02, 37.04, 
               36.79, 33.7, 32.8, 38.55, 37.74, 39.52, 33.69, 34.3, 33.89,
               34.28, 33.91, 40.88),
  age = c(30, 28, 34, 15, 43, 26, 18, 17, 15, 18, 26, 52, 48, 25, 17, 
          25, 16, 30, 31, 33),
  rooms = c(2948, 4225, 1748, 1701, 1584, 990, 1634, 4977, 2524, 2135, 
            2442, 2037, 1186, 848, 89, 2197, 5003, 3214, 2583, 2235),
  bedrooms = c(827, 835, 284, 346, 477, 223, 274, 994, 451, 373, 659, 
               358, 213, 153, 19, 320, 1180, 513, 663, 506),
  population = c(1635, 1908, 783, 723, 799, 719, 899, 1987, 1207, 1464, 
                 1134, 811, 487, 436, 79, 934, 4145, 1700, 1675, 1165),
  households = c(750, 686, 303, 352, 433, 232, 285, 947, 424, 405, 624, 
                 375, 207, 155, 21, 330, 1159, 533, 612, 441)
)
```

### Part a (2 points)

Using the training set, fit a linear regression model predicting house value from the other variables.

```{r}
house_model = lm(house_value ~ longitude + latitude + age + rooms + bedrooms + population + households, data = housing_train)
house_predict = predict(house_model, newdata = housing_train)
```

### Part b (4 points)

Fit a ridge regression using `cv.glmnet()` to choose the optimal $\lambda$ using "lambda.min". Set the seed before using`cv.glmnet()`.

```{r}
set.seed(0401)
w = model.matrix(house_value ~ longitude + latitude + age + rooms + bedrooms + population + households, data = housing_train)[, -1]
z = housing_train$house_value

cv_house = cv.glmnet(w, z, alpha = 0)
la = cv_house$lambda.min
ridge_predict = predict(cv_house, newx = w, s = 'lambda.1se')

```

### Part c (5 points)

Compute the sum of squared errors (that is $\sum_{i=1}^n (y_i-\hat{y}_i)^2$ for both the OLS model and the ridge regression model for the **training set**. The `predict()` functions will be useful here. Compare the results and comment on why the OLS performs better here (the OLS should indeed perform better).

```{r}
house_sse = sum((z - house_predict)^2)
house_sse

ridge_sse = sum((z - ridge_predict)^2)
ridge_sse

# The OLS has the smaller sum of errors. Ridge repression adds a penalty to shrink variables closer to zero, but this causes a sacrifice in the training error, which is why this is the case. 
```

### Part d (3 points)

Compute the sum of squared errors (that is $\sum_{i=1}^n (y_i-\hat{y}_i)^2$ for both the OLS model and the ridge regression model for the **testing set**. Remember to use the exact models you fit to the training sets. The `predict()` functions will be useful here. Compare the results and comment on why the ridge regression performs better here.

```{r}
set.seed(3939)
house_predict2 = predict(house_model, newdata = housing_test)
house_sse2 = sum((housing_test$house_value - house_predict2)^2)
house_sse2

m = model.matrix(house_value ~ longitude + latitude + age + rooms + bedrooms + population + households, data = housing_test)[, -1]
t = housing_test$house_value 

house_cv_ridge = cv.glmnet(w, z, alpha = 0)
ridge_predict2 = predict(house_cv_ridge, newx = m, s = "lambda.1se")
ridge_sse2 = sum((t - ridge_predict2)^2)
ridge_sse2

# Normally, the ridge regression has lower error, but in this case, it is slightly higher. The ridge regression usually performs better because it reduces overfitting in the model. However, this is not always the case. 
# ridge_sse2: 114660741972
# house_sse2: 111268808003
```

### Part e (7 points)

Use the `train()` function in the `caret()` package to find the optimal $\alpha, \lambda$ pair in this situation (using the training set and **10-fold cross validation**). For $\lambda$, search for values between 0 and 10000 by 100 and for $\alpha$, search for values between 0 and 1 by 0.05. Then fit a model using the optimal $\alpha$ and $\lambda$ using `glmnet()` and compare the sum of squared errors using this optimized model on the test set to the two models in part (d). Set a seed before running the `train()` function.

```{r}
#| cache: true
set.seed(3939)
library(caret)

alpha_grid = seq(0, 1, by = 0.05)
lambda_grid = seq(0, 10000, by = 100)

tune_grid = expand.grid(
  alpha = alpha_grid,
  lambda = lambda_grid
)

ctrl = trainControl(method = "cv", number = 10)

cv_model = train(
  x = w,                 # training predictors (matrix)
  y = z,                 # training response
  method = "glmnet",
  trControl = ctrl,
  tuneGrid = tune_grid
)

cv_model$bestTune

best_alpha = cv_model$bestTune$alpha
best_lambda = cv_model$bestTune$lambda

final_model = glmnet(w, z, alpha = best_alpha, lambda = best_lambda)

final_pred = predict(final_model, newx = m)

final_sse = sum((t - final_pred)^2)

final_sse
# The final_sse was lower than both the previous sse's used. 
# final_sse: 110599450410
```

# Problem 3 (41 points)

In the `AER` package is the data set "ShipAccidents". You can install that package and load that data using `data(ShipAccidents)`. Type `?ShipAccidents` to read about that data set.

### Part a (4 points)

Load in the data, remove the rows for which service is equal to 0, and then fit a Poisson regression model for predicting incidents from all other variables using the ShipAccidents data. Don't forget to put `family = "poisson"` in the `glm()` function. Briefly explain why it makes sense to do Poisson regression here.

```{r}
library(AER)
data(ShipAccidents)
ship_data = ShipAccidents[ShipAccidents$service != 0, ]
poisson_model = glm(incidents ~ ., data = ship_data, family = "poisson")
summary(poisson_model)
# It makes sense to use Poisson here, because we are measuring occurrences over a period of time.
```

### Part b (5 points)

Using that model, interpret the slope of "service" and the slope of "construction1970-74" in original (not log) units.

```{r}
coef(poisson_model)
exp(coef(poisson_model)["service"])
exp(coef(poisson_model)["construction1970-74"])
# Each additional unit of service increases the expected number of incidents by 0.0064%.
# Ships constructed in 1970-74 have an expected number of incidents that is 4.24 times that of ships in the reference construction category when all other variables are constant. 
```

### Part c (6 points)

Make a residual plot of the Pearson residuals vs the linear predictors and make a QQ plot of the Pearson residuals. Comment on what these imply about the fit.

```{r}
pearson_res = residuals(poisson_model, type = "pearson")
lin_pred = predict(poisson_model, type = "link")
plot(lin_pred, pearson_res,
     xlab = "Linear Predictor",
     ylab = "Pearson Residuals",
     main = "Pearson Residuals vs Linear Predictor")

abline(h = 0, col = "turquoise")

qqnorm(pearson_res, main = "QQ Plot of Pearson Residuals")
qqline(pearson_res, col = "red")
# The QQ plot seems to have heavy tails, making the data skewed right. It implies that the distribution is not a good fit. 
```

### Part d (5 points)

Conduct a deviance goodness-of-fit (GOF) test for a lack of fit here. Type out the null and alternative hypotheses, report the test statistic, give the p-value, and provide an interpretation.

```{r}
summary(poisson_model)
p_value = 1 - pchisq(poisson_model$deviance, poisson_model$df.residual)
p_value

# H0: The Poisson regression model fits the data well
# H1: The Poisson regression model does not fit the data well
# Test stat: D = 77.45 w/ 24 degrees of freedom

# p-value = 1.542447e-07
# Since the p-value is very small, we reject the null hypothesis. Thus, there is evidence that the model does not fit the data well. 

```

### Part e (4 points)

Is there evidence that this model has overdispersion? Explain.

```{r}
dispersion = sum(residuals(poisson_model, type = "pearson")^2) / poisson_model$df.residual
dispersion
# returns 2.896271, which is much greater than 1, so there is evidence of overdispersion. This means it is likely the variance is greater than the mean, which works against the assumptions we made in the using a Poisson distribution. 

```

### Part f (6 points)

Fit a "quasipoisson" model to these data. Using this model, obtain an approximate 95% confidence interval for the mean response for a ship of type "B" with construction between 1965 and 1969, with operation between 1960 and 1974, and with 2000 aggregate months of service. You can construct this interval using a t critical value with 24 degrees of freedom since the quasipoisson family more closely follows a t distribution than a normal one. Interpret the interval.

```{r}
quasi_model = glm(incidents ~ ., 
                  data = ship_data, 
                  family = "quasipoisson")
new_ship = data.frame(
  type = factor("B", levels = levels(ship_data$type)),
  construction = factor("1965-69", levels = levels(ship_data$construction)),
  operation = factor("1960-74", levels = levels(ship_data$operation)),
  service = 2000
)

pred = predict(quasi_model, newdata = new_ship, se.fit = TRUE, type = "link")

t_crit = qt(0.975, df = 24)

lower_log = pred$fit - t_crit * pred$se.fit
upper_log = pred$fit + t_crit * pred$se.fit

lower = exp(lower_log)
upper = exp(upper_log)

c(lower, upper)
# lower: 3.66
# upper: 19.95
# We can be 95% confident that the mean number of incidents of ships of type B constructed between 1965-69 that operated between 1960-74 with 2000 months of serves is between 3.66 and 19.95.
```

### Part g (5 points)

Now construct an approximate 95% prediction interval for the number of incidents for the ship described in part f using the `qpois()` function. Interpret the interval.

```{r}
pred = predict(quasi_model, newdata = new_ship, type = "response")
lambda_hat = pred
lower_pred = qpois(0.025, lambda = lambda_hat)
upper_pred = qpois(0.975, lambda = lambda_hat)

c(lower_pred, upper_pred)
# lower: 3
# upper: 15
# We can be 95% confident that the mean number of incidents of ships of type B constructed between 1965-69 that operated between 1960-74 with 2000 months of serves is between 3 and 15.
```

### Part h (6 points)

Suppose we did not catch the fact that this model is overdispersed. Repeat parts f and g using the model fit with the "poisson" family, not "quasipoisson" and using a z interval for the mean response rather than a t interval. Compare your results to parts f and g and comment on what is different.

```{r}
pred_pois = predict(poisson_model, newdata = new_ship, 
                    se.fit = TRUE, type = "link")
z_crit = 1.96

lower_log = pred_pois$fit - z_crit * pred_pois$se.fit
upper_log = pred_pois$fit + z_crit * pred_pois$se.fit

lower_pois = exp(lower_log)
upper_pois = exp(upper_log)

c(lower_pois, upper_pois)

lambda_hat_pois = predict(poisson_model, newdata = new_ship, type = "response")

lower_pred_pois = qpois(0.025, lambda = lambda_hat_pois)
upper_pred_pois = qpois(0.975, lambda = lambda_hat_pois)

c(lower_pred_pois, upper_pred_pois)
# Conf. Int: 5.33 - 13.71
# Prediction Int: 3 - 15
# The Poisson Confidence Interval has a much narrower range than the quasipoisson interval. However, the prediction interval is the same as the one we predicted. 
```
